---
title: "Homework 11"
author: "Richard Fischer"
date: "December 4, 2019"
output: html_document
---

```{r}
library(ggplot2)
require(data.table)

paydata <- as.data.frame(fread("State_Employee_Payroll_Data_Calendar_Year_2015_through_Present.tsv"))

paydata <- subset(paydata, paydata$`Full/Part` == "F")
attach(paydata)
paydata$fullName <- paste(`First Name`, `Middle Initial`, `Last Name`, sep = '')

df2 <- aggregate(`Bi-Weekly Comp Rate`, by = c(list(paydata$fullName),
                                              list(`Pyrl Fiscal Yr`)),
                                              median)
colnames(df2) <- c('Name', "Year", "Amount")

y2015 <- subset(df2, df2$Year == 2015)
y2015 <- cbind(y2015$Name, as.numeric(y2015$Amount))
colnames(y2015) <- c("Name", "2015")

y2016 <- subset(df2, df2$Year == 2016)
y2016 <- cbind(y2016$Name, as.numeric(y2016$Amount))
colnames(y2016) <- c("Name", "2016")

y2017 <- subset(df2, df2$Year == 2017)
y2017 <- cbind(y2017$Name, as.numeric(y2017$Amount))
colnames(y2017) <- c("Name", "2017")

y2018 <- subset(df2, df2$Year == 2018)
y2018 <- cbind(y2018$Name, as.numeric(y2018$Amount))
colnames(y2018) <- c("Name", "2018")

y2019 <- subset(df2, df2$Year == 2019)
y2019 <- cbind(y2019$Name, as.numeric(y2019$Amount))
colnames(y2019) <- c("Name", "2019")

y2020 <- subset(df2, df2$Year == 2020)
y2020 <- cbind(y2020$Name, as.numeric(y2020$Amount))
colnames(y2020) <- c("Name", "2020")


newdata <- merge(y2015, y2016, by.x = 'Name', by.y = 'Name', all.x = TRUE, all.y = TRUE)
newdata <- merge(newdata, y2017, by.x = 'Name', by.y = 'Name', all.x = TRUE, all.y = TRUE)
newdata <- merge(newdata, y2018, by.x = 'Name', by.y = 'Name', all.x = TRUE, all.y = TRUE)
newdata <- merge(newdata, y2019, by.x = 'Name', by.y = 'Name', all.x = TRUE, all.y = TRUE)
newdata <- merge(newdata, y2020, by.x = 'Name', by.y = 'Name', all.x = TRUE, all.y = TRUE)

newdata[newdata == 0] <- NA

newdata$`2015` <- as.numeric(newdata$`2015`)
newdata$`2016` <- as.numeric(newdata$`2016`)
newdata$`2017` <- as.numeric(newdata$`2017`)
newdata$`2018` <- as.numeric(newdata$`2018`)
newdata$`2019` <- as.numeric(newdata$`2019`)
newdata$`2020` <- as.numeric(newdata$`2020`)


newdata$raise2016 <- (newdata$`2016` - newdata$`2015`) / newdata$`2015`
newdata$raise2017 <- (newdata$`2017` - newdata$`2016`) / newdata$`2016`
newdata$raise2018 <- (newdata$`2018` - newdata$`2017`) / newdata$`2017`
newdata$raise2019 <- (newdata$`2019` - newdata$`2018`) / newdata$`2018`
newdata$raise2020 <- (newdata$`2020` - newdata$`2019`) / newdata$`2019`

head(newdata)
```

For this dataset, I had to do a lot of data wrangling. First, I subsetted the data to only include full time, regular employees. Then, I concatenated the first, middle, and last name together to get a set of full names, and found the median bi-weekly payment for each person for each fiscal year. After that, I transformed this into a matrix with each row representing each person and each column representing the median bi-weekly payment for a year. Finally, I found the percentage difference between each column and the column to the left of it to calculate the percentage raise that each person received each year.

```{r}
newdata <- newdata[, c('Name', 'raise2016', 'raise2017', 'raise2018', 'raise2019', 'raise2020')]
newdata <- data.table(newdata)
raises <- melt.data.table(newdata, 
                          id.vars = c("Name"), 
                          measure.vars = c("raise2016", "raise2017", "raise2018", "raise2019", "raise2020"),
                          na.rm = TRUE)

ggplot(raises, aes(x=variable, y=value)) + 
  geom_boxplot() +
  scale_y_continuous(limits = quantile(raises$value, c(0.1, 0.9)))
```

The first figure I made is a boxplot of all of the raises by year. We see that 2016 had a lot of raises, but then the raises decreased to an all time low in 2018. Then, they came back up by 2020.

```{r}
ggplot(raises) +
  geom_histogram(aes(value))
```

Here I wanted to look at the distribution of the raises. Because of some outliers, our scale is very off, so I changed the x limit for the chart.

```{r}
ggplot(raises) +
  geom_histogram(aes(value)) +
  xlim(-0.4, 0.8)
```

Here we see that a lot of people are getting small raises, and larger raises steadily become less common. Also, there are a lot of times when people are getting paid less than the year before.

```{r}
agedata <- aggregate(Age, by = c(list(paydata$fullName)),
                                              min)
colnames(agedata) <- c('Name', "Age")
raisesbyAge <- merge(raises, agedata, by.x = "Name", by.y = "Name")

raisesbyAge <- aggregate(raisesbyAge$value, by = c(list(raisesbyAge$Age)), median)
raisesbyAge <- raisesbyAge[2:nrow(raisesbyAge), ]
colnames(raisesbyAge) <- c("Age", "Median_Raise")

ggplot(raisesbyAge, aes(x = Age, y = Median_Raise)) + 
  geom_col()
```

Here I made a plot of the median raise for each age group. It appears that raises are larger for younger people.

```{r}
cor.test(raisesbyAge$Age, raisesbyAge$Median_Raise)
```

Here I wanted to confirm that there is a correlation between age and the size of the raise, so I did a correlation test. We can see that since our p-value is still relatively high, we cannot assume that there is a relationship between age and raise amounts.

```{r}
egdata <- aggregate(`Bi-Weekly Comp Rate`, by = c(list(paydata$fullName), list(`Ethnic Grp`)), median)
colnames(egdata) <- c('Name', "Ethnic_Group", "Payment")
raisesbyEG <- merge(raises, egdata, by.x = "Name", by.y = "Name")
raisesbyEG <- aggregate(raisesbyEG$value, by = c(list(raisesbyEG$Ethnic_Group)), median)
colnames(raisesbyEG) <- c("Ethnic_Group", "Median_Raise")

ggplot(raisesbyEG, aes(x = Ethnic_Group, y = Median_Raise)) + 
  geom_col()
```

Here we see the median raise percentage by ethnic group.

```{r}
genderdata <- aggregate(`Bi-Weekly Comp Rate`, by = c(list(paydata$fullName), list(Sex)), median)
colnames(genderdata) <- c('Name', "Gender", "Payment")
raisesbyGender <- merge(raises, genderdata, by.x = "Name", by.y = "Name")

ggplot(raisesbyGender, aes(x=Gender, y=value)) + 
  geom_boxplot() +
  scale_y_continuous(limits = quantile(raisesbyGender$value, c(0.1, 0.9)))
```

Here I made a boxplot of the raise percentage by gender. By looking at this, it appears that there isn’t any difference in the size of the raise based on a person’s gender.

```{r}
summary(aov(raisesbyGender$value ~ raisesbyGender$Gender))
```

```{r}
aggregate(raisesbyGender$value, by = c(list(raisesbyGender$Gender)), mean)
```

I tried to confirm this with an anova test, and we see that since our p-value is low, we can actually assume that there is a different mean raise percentage for different genders. I also calculated the sample mean for each gender group, and I found that the U category had a much higher mean. To compare only male and females, I decided to exclude the U category and do the test again.

```{r}
raisesbyGender <- subset(raisesbyGender, 
                         raisesbyGender$Gender != "U")
summary(aov(raisesbyGender$value ~ raisesbyGender$Gender))
```

When I did the test with just male vs female, I got a high p-value, so I concluded that there is no difference between the amount of raises that men get compared to women.





